\section*{Solution}

%    \lstset{basicstyle=\scriptsize\sf}
%    \lstinputlisting{./src/ex1/main.cpp}
%    \lstset{basicstyle=\sf}

The programs which use the traits are reported below

\begin{enumerate}

    \item An implementation is the following:
    \lstset{basicstyle=\scriptsize\sf}
    \lstinputlisting{./src/solution/isSame/main.cpp}
    \lstset{basicstyle=\sf}
    First of all we need to include the \cpp{<type_traits>} to have the access to
    the methods to handle the types. This functionality is introdued from the C++11.
    We can access to different traits and, in particular for our case, to the class
    \cpp{std::is_same}. The latter takes in input two template parameters needed for
    the comparison, we can access to its member \cpp{value} to determine if they
    are equal or not.  This operation is done statically. After the evaluation of
    both the \cpp{std::is_same} we have two boolean linked with the \cpp{||}
    operator.  Also in this case the operation is done statically, since the
    values
    are known.  Finally we can use the \cpp{static_assert} to rise a compilation
    error if the condition, specified in the first argument, is \cpp{false}.
    The \cpp{static_assert} is checked only during the compilation and not at
    run time. \\
    Compiling the program we obtain
    \begin{verbatim}
main.cpp:7:7: error: static_assert failed "Bad value"
    { static_assert( std::is_same<T,int>::value ||
std::is_same<T,double>::value, "Bad value" ); }
      ^              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
main.cpp:15:13: note: in instantiation of member function 'B<char>::B' requested
here
    B<char> a;
            ^
1 error generated.
    \end{verbatim}
    The template parameter \cpp{char} is not allowed by the class.

    \item An implementation is the following:
    \lstset{basicstyle=\scriptsize\sf}
    \lstinputlisting{./src/solution/pointer/main.cpp}
    \lstset{basicstyle=\sf}
    Since the input parameter of \cpp{printValue} is a-priori unknown we have to
    check if it is a raw pointer or not. We use the class \cpp{std::is_pointer}
    to check if the type, specified as the first template parameter, is a
    pointer or not.
    Beware, this class does not work as you expect with \cpp{std::shared_ptr}
    but it works only with raw pointers.

    \item An implementation is the following:
    \lstset{basicstyle=\scriptsize\sf}
    \lstinputlisting{./src/solution/conditional/main.cpp}
    \lstset{basicstyle=\sf}
    We need a method that can do a conditional chose depending on a boolean.
    In particular the return value have to be a type. The most easy way is to
    use \cpp{std::conditional}, which takes three template paramters: the
    boolean as first then the two possible types. In our case if \cpp{check}
    is \cpp{true} then \cpp{new_type} is a \cpp{float} otherwise is a \cpp{int}.
    To deduce the type of the variable \cpp{i} we use the function
    \cpp{decltype} then, using the trait \cpp{std::is_signed}, we can understand
    if \cpp{i} is a signed or unsigned type.

\end{enumerate}

To construct the dot product using the template metaprogramming the key
ingredients are templates and \cpp{static} and \cpp{inline} functions. In this way the
compiler can perform a lot of optimization. First of all we need a class,
templetized on the index of the array, that perform the operations.
Its implementation is reported below.
\lstset{basicstyle=\scriptsize\sf}
\lstinputlisting{./src/solution/MetaDot/metadot.hpp}
\lstset{basicstyle=\sf}
As the comment says the class is templetized on the \cpp{std::size_t}, since it
returns the correct type used to dimension array. This type it can be
implementation dependent (32 or 64 bits) it is better to use it instead of just
unsigned int. The class \cpp{metaDot} contains the static method \cpp{apply}
which is templetize on the same template variable of the \cpp{std::array}.
Notice that we statically check that the two arrays have the same dimension. The
method simply perform a single operation of the dot product and then call again
\cpp{apply} with the index decreased by one. This strategy lead us to perform
the dot product recursively. The method is static, so we do not need an object
of type \cpp{metaDot} for each of the index.  Moreover the implementation is
done inside the class so the inlining is automatic.  Finally we need a stopping
criteria, so we implement a specialization of \cpp{apply} for the first index.
The free function \cpp{operator *} mask the call of the method \cpp{apply} by
using the normal \cpp{*} operator. The return value is a \cpp{constexpr T} which
can be more usefull than only the value.\\
The use of the dot product is reported below.
\lstset{basicstyle=\scriptsize\sf}
\lstinputlisting{./src/solution/MetaDot/main_metadot.cpp}
\lstset{basicstyle=\sf}
