\documentclass[10pt]{beamer}
\usetheme{default}
\setbeamercovered{invisible}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}{
    \flushright{\hfill \insertframenumber{}/\inserttotalframenumber}
}

\usepackage{listings}

% User-defined colors.
\definecolor{DarkGreen}{rgb}{0, .5, 0}
\definecolor{DarkBlue}{rgb}{0, 0, .5}
\definecolor{DarkRed}{rgb}{.5, 0, 0}
\definecolor{LightGray}{rgb}{.95, .95, .95}
\definecolor{White}{rgb}{1.0,1.0,1.0}
\definecolor{darkblue}{rgb}{0,0,0.9}
\definecolor{darkred}{rgb}{0.8,0,0}
\definecolor{darkgreen}{rgb}{0.0,0.85,0}

% Settings for listing class.
\lstset{
  language=C++,                        % The default language
  basicstyle=\small\ttfamily,          % The basic style
  backgroundcolor=\color{White},       % Set listing background
  keywordstyle=\color{DarkBlue}\bfseries, % Set keyword style
  commentstyle=\color{DarkGreen}\itshape, % Set comment style
  stringstyle=\color{DarkRed}, % Set string constant style
  extendedchars=true % Allow extended characters
  breaklines=true,
  basewidth={0.5em,0.4em},
  fontadjust=true,
  linewidth=\textwidth,
  breakatwhitespace=true,
  showstringspaces=false,
  lineskip=0ex, %  frame=single
}

\begin{document}
    \title{A Runge-Kutta-Fehlberg solver\protect\\using traits and \texttt{Eigen}\protect\\(part I)}
    \author{Pasquale Claudio Africa}
    \date{08/04/2022}

\begin{frame}[plain, noframenumbering]
    \maketitle
\end{frame}

\begin{frame}{RKF solver}
This exercise (a modified version of \texttt{Examples/src/RKFSolver}) is about a set of tools that implements embedded Runge-Kutta-Fehlberg methods to solve non-linear scalar and vector Ordinary Differential Equations, based on the \texttt{Eigen} library.
\vfill
Consider an ODE of the form
\begin{equation*}
\frac{\mathrm{d} y}{\mathrm{d} t} = f(t, y),
\end{equation*}
and consider the coefficients \(b_i, b_i^*, a_{ij}, c_i\) to be given in the form of a Butcher tableau.
We recall that, at each time step \(t_n\), an embedded RKF method of order \(p\) consists of the following steps.
\begin{enumerate}
\item Compute the high-order solution \(y_{n+1} = y_{n} + h_n \sum_{i=1}^p b_i k_i\), where \(k_i = f\left(t_n + c_i h_n, y_n + h_n\sum_{j=1}^{i-1}a_{ij}k_j\right)\), and \(h_n = t_{n+1} - t_n\).
\item Compute the low-order solution \(y_{n+1}^* = y_{n} + h_n \sum_{i=1}^p b_i^* k_i\).
\item Compute the error \(\varepsilon_{n+1} = y_{n+1} - y_{n+1}^* = h_n \sum_{i=1}^p(b_i - b_i^*)k_i\).
\item Adapt the step size \(h_{n+1} = \tau_{n} h_{n}\), where \(\tau_{n}\) is a prescribed reduction (\(<1\))/expansion (\(>1\)) factor depending on whether \(\varepsilon_{n+1}\) is larger or smaller than a prescribed tolerance.
\end{enumerate}
\end{frame}

\begin{frame}{Exercise: RKF solver}
The code structure is the following:
\begin{itemize}
\item \texttt{ButcherRKF} contains the definition of the Butcher tableaux for some common RKF methods.
\item \texttt{RKFTraits} defines the basic structure that enables to statically select the type of the equation(s) to be solved (\textit{i.e.} scalar or vector).
\item \texttt{RKFResult} is a data structure containing the output of the RKF solver.
\item \texttt{RKF} implements a generic RKF solver interface, filling a proper \texttt{RKFResult} object.
\end{itemize}
\vfill
Starting from the provided solution sketch:
\begin{enumerate}
\item Implement the \texttt{RKFTraits} class defining common types for homogeneity.
\item Implement the \texttt{RKF::RKFstep} method, performing just one timestep of the RKF method.
\item Implement the \texttt{RKF::operator()} time-advancing method, without error correction.
\item Use the just implemented solver to solve the model problem and the Van der Pol oscillator problem defined in the \texttt{main.cpp} file.
\item Include error correction into the solver and compare the results.
\end{enumerate}
\end{frame}

\end{document}
